---
title: "SimulatedAnalysis"
author: "Reed Scott"
date: "2025-02-06"
output: html_document
---

# Introduction

Hello Brittany!

I have been working on some simulated analysis for VTADs. The goal was to simulate data similar to what we're trying to do 
with VTADs, so that when we finally get results we can almost immediately plug and play. This does require some discussion of
what our hypotheses will be, which I discuss in an
earlier doc I sent (Hypotheses.docx). I think that that will help inform the full model set we will build for this project
but for right now I wanted to create a simulated dataset and compare a few simple models.

I have choosen to simulate a single season occupancy model, and I have choosen to work in unmarked. My reasoning for a single
season is that for our real data, while we have multiple seasons it's only two years and seperated by a missing year of data.
So, I am suggesting we run 2 single season models. I want to work in unmarked because 1) I think it's good professional
development to learn the multiple ways of doing ecological modeling and 2) I think unmarked is a bit easier to work with for
smaller model sets.

With all this in mind, let's look at the simulation I did

# Building the simulated data

The first thing I wanted to do was set the parameters that would define our data. In this simulation I chose to use 24 ponds
as our number of sites. I simulated 3 visits, and 3 tests. However, unmarked couldn't handle 108 tests (12 * 3 * 3) for
simulation purposes, so instead of having 12 frogs per survey I only had 2 (so 18 total tests). With these paramters in place
we can form our detection matrix, which ends up being 24x18 like we would expect.

```{r Init params + det matrix}
#Initialize -----------------------------------------
rm(list=ls())
library(unmarked)
library(MuMIn)
library(ggplot2)
# Load functions--------------------------------------


# Global Variables-------------------------------------


# Program Body------------------------------------------
# Initialize variables
set.seed(1234)
ponds <- 24 #number of sites
frogs <- 2 #number of frogs PER VISIT
visits <- 3 #number of visits to each site
tests <- 3 #number of tests per frog
total_frogs <- frogs*visits
total_tests <- frogs*visits*tests #total number of tests for each pond for the season

# Create an empty dataframe for the detection data
det <- matrix(NA, nrow = ponds, ncol = total_tests)
```

Now the goal is to populate this empty dataframe based on covariates that effect $\psi$, $\theta$, and p. To do this in unmarked you need to
create seperate objects for the site covariates, availability covariates, and observation covariates. I am going to spare you the code for the
moment, but I did want to take this opporunity to reiterate my hypotheses at the moment. I have hypotheses that $\psi$ will be negatively 
correlated with $\beta$ and $\alpha$, and positively correlated with $R_{0P}$. For $\theta$ I believe the relationships with $\beta$, $\alpha$,
and $R_{0P}$ to be the same, but adding in an effect of Host Species ID. The hypotheses for p are all the same as for $\theta$.

```{r Covar setup, echo = FALSE}
site_covs <- data.frame(betadiv = rnorm(ponds),
                        alpha = rnorm(ponds),
                        R0P = rnorm(ponds))
Spp <-  data.frame(matrix(sample(c("Spp1","Spp2","Spp3","Spp4"), size = total_frogs*ponds, replace = T), nrow = ponds, ncol = total_frogs))
Spp <-  data.frame(lapply(Spp,as.factor))

avail_covs <- list(betadiv = data.frame(data = matrix(rep(site_covs$betadiv, each = total_frogs), nrow = ponds, ncol = total_frogs, byrow = T)),
                   alpha = data.frame(data = matrix(rep(site_covs$alpha, each = total_frogs), nrow = ponds, ncol = total_frogs, byrow = T)),
                   R0P = data.frame(data = matrix(rep(site_covs$R0P, each = total_frogs), nrow = ponds, ncol = total_frogs, byrow = T)),
                   Spp = Spp) 
names(avail_covs$R0P) <- c("Ind1","Ind2","Ind3","Ind4","Ind5","Ind6")
Spp_dets <- NULL
for(i in 1:ncol(Spp)){
  Spp_temp <- matrix(rep(Spp[,i], each = tests), nrow = ponds, ncol = tests, byrow = T)
  Spp_dets <- cbind(Spp_dets, Spp_temp)
}
Spp_dets <- data.frame(lapply(data.frame(Spp_dets),as.factor))    
    
test_covs <- list(betadiv = data.frame(data = matrix(rep(site_covs$betadiv, each = total_tests), nrow = ponds, ncol = total_tests, byrow = T)),
                  alpha = data.frame(data = matrix(rep(site_covs$alpha, each = total_tests), nrow = ponds, ncol = total_tests, byrow = T)),
                  R0P = data.frame(data = matrix(rep(site_covs$R0P, each = total_tests), nrow = ponds, ncol = total_tests, byrow = T)),
                  Spp = Spp_dets)
for(i in 1:length(test_covs)){
  for (j in 1:ncol(test_covs$betadiv)){
    names(test_covs[[i]])[[j]] <- paste("test",i, sep = "")
  }
}
```

With the covariates simulated, we're now ready to create and unmarkedFrame object and simulate detection data based on our covariates. In this example, I simulated a negative effect of $\beta$ on $\psi$, a Species effect on $\theta$, and an effect of $\alpha$ on p. 

```{r Make umf + sim data}
umf <- unmarkedFrameGOccu(y = det, siteCovs = site_covs, obsCovs = test_covs, yearlySiteCovs = avail_covs, numPrimary = total_frogs)
coeff <- list(psi = c(0,-5), 
              phi = c(0,-0.9,0.8,0.5), 
              det = c(0,0.75))
sim2 <- simulate(umf, model = goccu,  psiformula = ~betadiv, phiformula = ~Spp, pformula =~ alpha, coefs = coeff)
```

And just like that we've simulated some data! But does it work? We can check using the goccu function. In a simple example below, I fit a null model and three models for the variables effecting $\psi$. If the simulation worked correctly, then psi.beta should be the top model.

```{r Create + fit models}
#null model
null.mod <- goccu(psiformula = ~1, phiformula = ~1, pformula = ~1,data = sim2[[1]])
#Models for psi
psi.beta <- goccu(psiformula = ~betadiv, phiformula = ~1, pformula = ~1, data = sim2[[1]])
psi.alpha <- goccu(psiformula = ~alpha, phiformula = ~1, pformula =~ 1, data = sim2[[1]])
psi.R0P <- goccu(psiformula = ~R0P, phiformula = ~1, pformula =~ 1, data = sim2[[1]])

fitlist.psi <- fitList("nullpsi" = null.mod,
                   "betapsi" = psi.beta,
                   "alphapsi" = psi.alpha,
                   "R0Ppsi" = psi.R0P)

modSel(fitlist.psi)
```

And it is! 

We can check the same thing for $\theta$, and this should show that the top model includes an affect of species:

```{r Theta mods}
#Models for theta
theta.beta <- goccu(psiformula = ~1, phiformula = ~betadiv, pformula = ~1, data = sim2[[1]])
theta.alpha <- goccu(psiformula = ~1, phiformula = ~alpha, pformula = ~1, data = sim2[[1]])
theta.R0P <- goccu(psiformula = ~1, phiformula = ~R0P, pformula = ~1, data = sim2[[1]])
theta.spp <- goccu(psiformula = ~1, phiformula = ~Spp, pformula = ~1, data = sim2[[1]])

fitlist.theta <- fitList("beta" = theta.beta,
                         "alpha" = theta.alpha,
                         "R0P" = theta.R0P,
                         "Spp" = theta.spp)
modSel(fitlist.theta)
```

This indicates that the simulation worked and the models are behaving as expected.

# Conclusion + next steps

I think this indicates that we can feasibly use unmarked for VTADs. I think that we need to have a discussion about what our hypotheses are
and then this will inform what models we would like to test.